---
title: "Filematrix overview"
author: "Andrey Shabalin, AAAAAAAAAAAAAAHHHHHH"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true # table of content true
vignette: >
  %\VignetteIndexEntry{Filematrix overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Using the class `filematrix` one can create and access large matrices stored in files, not computer memory.
Filematrices can be as large as the storage allows. The package has been tested on matrices several terabytes in size.


### Creating filematrices

File matrices can be created using functions `fm.create`, `fm.create.from.matrix`, or `fm.create.from.text.file`:

```{r}
suppressPackageStartupMessages( library(filematrix) )
fm = fm.create(filenamebase = 'fmatrix', nrow = 1000, ncol = 1000, type = 'double')
```

The code above creates two files: `fmatrix.bmat` which stores the filematrix values and `fmatrix.desc.txt` which stores the filematrix information, such as filematrix dimensions, data type, and data type size.

### Reading from and writing to a filematrix

The elements of a filematrix can be read and changed using the same syntaxis as used for regular R matrices.
Any changes to a filematrix are **immediately** written to the `.bmat` file.


```{r}
fm[1:3, 1:2] = 1:6;
fm[1:4, 1:3]
colSums(fm[,1:4])
```

Elements of a filematrix can also be accessed as elements of a vector. Note that `fm` has 1000 rows, so `fm[1,2]` accesses the same as `fm[1001]`.

```{r}
fm[1:4]
fm[1001:1004]
```

### Row and column names

File matrices can also have row and column names, like regular R matrices.

```{r}
colnames(fm) = paste0('Col',1:ncol(fm))
rownames(fm) = paste0('Row',1:nrow(fm))
```

The row and column names of the filematrix `fm` are stored in `fmatrix.nmsrow.txt` and `fmatrix.nmscol.txt` respectively.

### Closing filematrices

An open filematrix object can be closed with `close` function. This closes the internal file handle (connection).
Closing filematrix objects is **optional**, changes **would not be lost** if the object is not closed.

```{r}
close(fm)
```

### Open or load an existing filematrix

An existing filematrix can be opened with `fm.open`.

```{r}
fm = fm.open(filenamebase = 'fmatrix');
```

An existing filematrix that would fit in memory can be loaded fully with `fm.load`

```{r}
mat = fm.load('fmatrix');
```


### By column access is faster then by rows

The values of a filematrix are stored by columns, as with regular R matrices:

```{r}
matrix(1:12,3,4)
```

Thus, access to a filematrix values by columns is much faster than access by rows:

```{r}
timecol = system.time( { sum(fm[ ,1:10]) } )[3]
timerow = system.time( { sum(fm[1:10, ]) } )[3]
cat('Reading ', nrow(fm)*10, ' values from 10 columns takes ', timecol, ' seconds', '\n',
    'Reading ', ncol(fm)*10, ' values from 10 rows takes ',    timerow, ' seconds', '\n', sep = '')
```

### Appending columns

Unlike with regular R matrices, columns can be appended to the right side of a filematrix with very little overhead.

```{r}
dim(fm)
fm$appendColumns(nrow(fm):1)
fm$appendColumns(1:nrow(fm))
dim(fm)
fm[nrow(fm)+(-1:0), ncol(fm)+(-1:0)]

```

### Deleting filematrix files

If you no longer need a filematrix and want to delete its files from the hard drive and close the object,
please use `closeAndDeleteFiles()`

```{r}
closeAndDeleteFiles(fm)
```
